Classification of shapes

### 2019-08-06 manuscript/Manuscript2.Rnw
Manuscript update
updated manuscript

\subsection{Continuous Curves1}
Rather than describing a closed shape with a discrete set of points, a function can be used: 
$$
\beta:  \mathbb{S} \rightarrow \mathbb{R}^n 
$$
where $\mathbb{S}$ is the unit sphere and guarantees that the function has the same start and end point (i.e. is a closed curve).  The function $\beta$ is a closed curve but scalings, translations, rotations and reparameterizations of this function will usually result in different functions, which means that this framework does not offer invariance across these transformations.  Rather than using this function directly, the first derivative could be used, which results in a translation invariant representation.

This representation is the foundation of the square root velocity function (SRVF) (CITE) framework, which allows invariance across tranlsation, rotation, (optional) scaling, and reparameterization.  Frameworks for representing shapes that have invariance over reparameterization are said to be {\em elastic}.

@@

where $\dot{\beta(t)}$ is the first derivative of the function $\beta(t)$.

A full review of other methods for shape analysis can be found in CITE(FSDA)

@@

\section{Missing Data}
Missing data problems arise in many statistical settings due to survey non-response, designed missingness in surveys (CITE), measurement error, and latent classes, to name a few examples.  There are numerous methods proposed in the literaure describing methods for handling missing data in a statistically principled manner.  For example, (Demspter Laird and Rubin) describes the EM algorithm for finding maximum likelihood estimates in the presence of missing data.  Another method, of particular interest in this setting, for handling mising data is multiple imputation (CITE Rubin and Little, Schafer).  The idea behind multiple imputation is to blah blah blah.......

EM algorithm (Depster Laird and Rubin)
Multiple Imputation (Rubin and Little)

For a more thorough review of the missing data literature, see cite(HortonAndKleinman2007) (Add more).

\section{Missing Data and Shapes}

\subsection{13.1 - Incomplete Data}
Papers mentioned: 
\begin{itemize}
\item Albers and Gower (2010) - Missing data in procrustes
\item Mitchelson (2013) - handling occluded landmarks. 
\item Gunz et. al. (2009) - missing data in studies of biological evolution of bone surfaces
\end{itemize}

@@

Now for a fixed $j$, after computing $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)$ for all $i = 1, 2, \ldots, n_f$, the indices of the donor set are defined to be $D^{\star}_j = \{i : \Phi^{\star}_{\theta_j}(C_j^p,C_i^f) \le \Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(k)} \}$. $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(k)}$ is the $k$-th smallest value of $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(k)}$ over all $i = 1, 2, \ldots, n_f$ (e.g. $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(1)}$ and $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(n_p)}$ are the minimum and the maximum, respectively).

@@

\section{Simulation Study}
This simulation study is based on the shapes of the occlusal surfaces of of extant bovid molars both upper and lower.  Specifically, teeth from the following seven tribes: Alcelaphini, Antilopini, Bovini, Hippotragini, Neotragini, Reduncini, and Tragelaphini.  Specific counts of the sample sizes of each tooth type and tribe can be found in table \ref{sampsize}.

For each tooth, the raw data was made up of 60 points traced around the occlusal surface of the tooth.  For each tooth, these 60 points were split into two groups roughly divided by a line connecting the mesostyle to the entostyle/metastylid to the ectostylid of mandibular teeth/upper teeth.
An example of this can be seen in Figure \ref{ToothCutter}.  This cut was chosen for the simulation study and this is a break point commonly observed by in bovid fossils by biological anthropologists.

For each half of a tooth, the tooth was matched against all the complete teeth in the tooth type to find $k$ suitable donor teeth, the tooth was then completed $M$ times by randomly drawing $M$ teeth with replacement from the set of $k$ donor teeth, and using the donor tooth to complete the tooth as described in section XXX.

@@

complete_shape <- t(ptsTrainList[[1]][[1]])
plot(t(complete_shape), type = "l")
points(t(complete_shape), pch = 16)

@@

side <- 2
partial_shape <- t(tooth_cutter(ptsTrainList[[tooth]][[1]])[[side]])
points(t(partial_shape), pch = 16, col = "blue")
points(t(partial_shape), type = "l", col = "blue")
@

\caption{Example cuts for an Alcelaphini LM1 (DSCN2871)}
\label{ToothCutter}
\end{figure}

<<>>=
  k = 5
  M = 5
  side <- 1
  tooth <- "LM3"
load(paste0("/Users/gregorymatthews/Dropbox/shapeanalysisgit/results/results20190610_side=",side,"_k=",k,"_M=",M,"_tooth=",tooth,".RData"))

plot(t(results_list[[1]]$imputed_partial$imputed[[1]]),col = "red", type = "l", xlim = c(-200,700), ylim = c(-200,200))
points(t(results_list[[1]]$imputed_partial$partial_obs))
points(t(results_list[[1]]$imputed_partial$imputed[[2]]),col = "orange", type = "l")
points(t(results_list[[1]]$imputed_partial$imputed[[3]]),col = "gold", type = "l")
points(t(results_list[[1]]$imputed_partial$imputed[[4]]),col = "darkgreen", type = "l")
points(t(results_list[[1]]$imputed_partial$imputed[[5]]),col = "blue", type = "l")

greg <- c()
for (i in 1:length(results_list))
  {
  greg[i] <- results_list[[i]]$truth$Tribe
  }
  table(greg)
  sum(table(greg))
@

  \begin{table}[h!]
  \begin{center}
\begin{tabular}{ c c c c c c c c  }
  & Alcelaphini  & Antilopini & Bovini &Hippotragini & Neotragini & Reduncini & Tragelaphini & Total \\ 
 LM1 & 106 & 27 & 22 & 26 & 45 & 76 & 53 & 355 \\ 
 LM2 & 117 & 30 & 25 & 37 & 55 & 90 & 53 & 407\\ 
 LM3 & 117 & 30 & 19 & 34 & 53 & 96 & 62 & 411 \\ 
 UM1 & 117 & 30 & 23 & 37 & 43 & 80 & 75 & 405 \\ 
 UM2 & 118 & 30 & 23 & 41 & 53 & 97 & 94 & 456\\ 
 UM3 & 71 & 30 & 17 & 58 & 52 & 110 & 78 &  416\\ 
\end{tabular}
\end{center}
\caption{Sample sizes for the simulation study by tooth type and tribe}
\label{sampsize}
\end{table}

@@

>>>>

\subsection{Continuous Curves1}
Rather than describing a closed shape with a discrete set of points, a function can be used: 
$$
C:  \mathbb{S} \rightarrow \mathbb{R}^n 
$$
where $\mathbb{S}$ is the unit sphere and guarantees that the function has the same start and end point (i.e. is a closed curve).  The function $C$ is a closed curve but scalings, translations, rotations and reparameterizations of this function will usually result in different functions, which means that this framework does not offer invariance across these transformations.  Rather than using this function directly, the first derivative could be used, which results in a translation invariant representation.

This representation is the foundation of the square root velocity function (SRVF) \cite{FSDA} framework, which allows invariance across tranlsation, rotation, (optional) scaling, and reparameterization.  Frameworks for representing shapes that have invariance over reparameterization are said to be {\em elastic}.

@@

where $\dot{C(t)}$ is the first derivative of the function $C(t)$.

A full review of other methods for shape analysis can be found in \cite{FSDA}

@@

\section{Missing Data}

Missing data problems arise in many statistical settings due to survey non-response (CITE), designed missingness in surveys \cite{LittleRhemtulla2013,WoodEtAl2018}, measurement error (CITE), and latent classes (CITE), to name a few examples.  There are numerous methods proposed in the literaure describing methods for handling missing data in a statistically principled manner.  For example, (Demspter Laird and Rubin) describes the EM algorithm for finding maximum likelihood estimates in the presence of missing data.  Another method, of particular interest in this setting, for handling mising data is multiple imputation \cite{LittleRubin1987,SchaferBook}.  The idea behind multiple imputation is to contruct a model for the missing values conditional on the observed values, and then fill in the missing values by randomly drawing from this distribution.  This process is repeated $M$ times to create $M$ completed data sets.  An analysis of interest is performed on each of the $M$ completed data sets (e.g. calculating a mean, multiple regression, clustering, etc.) and then the results are combined across the completed data sets using appropriate combining rules \cite{LittleRubin1987}.

For a more thorough review of the missing data literature, see \cite{HortonKleinman2007) (MORE).  
\section{Missing Data and Shapes}
Much of the research involving missing data in shape analysis has focused on work related to landmark based data \cite{GunzEtAl2009,GunzEtAl2004,NeeserEtAl2009, CouetteAndWhite2010}. By defining landmarks on each shape, the problem can be framed in a more traditional statistical setting where each landmark can be thought of as a variables and more traditional missing data techniques can be used.  
%\cite{GunzEtAl2009} presented a method for reconstructing hominin crania using landmarks and semi-landmarks.

%\cite{GunzEtAl2004} Basically the same thing as the 2009 paper.

%\cite{NeeserEtAl2009} Also landmarks.  Using mean imputtion, thin plate spliens and regression.

%\cite{CouetteAndWhite2010}: Use mutivariate regression

\cite{Robinson2012}: Partial shape matching paper at Florida State.

\subsection{13.1 - Incomple e Data}
Papers mentioned: 
\begin{itemize}
\item Albers and Gower (2010) - Missing data in procrustes
\item Mitchelson (2013) - handling occluded landmarks. 
\item Gunz et. al. (2009) - missing data in studies of biological evolution of bone surfaces
\end{itemize}

@@

Now for a fixed $j$, after computing $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)$ for all $i = 1, 2, \ldots, n_f$, the indices of the donor set are defined to be $D^{\star}_j = \{i : \Phi^{\star}_{\theta_j}(C_j^p,C_i^f) \le \Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(k)} \}$. $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(k)}$ is the $k$-th smallest value of $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(k)}$ over all $i = 1, 2, \ldots, n_f$ (e.g. $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(1)}$ and $\Phi^{\star}_{\theta_{ij}}(C_j^p,C_i^f)_{(n_f)}$ are the minimum and the maximum, respectively).

@@

\section{Simulation Study}
This simulation study is based on the the shapes of the occlusal surface of upper and lower extant bovid molars.  Specifically, teeth from seven bovid tribes (Alcelaphini, Antilopini, Bovini, Hippotragini, Neotragini, Reduncini, and Tragelaphini) and twenty species (arundium, buselaphus, caffer, campestris, capreolus, dorcas, ellipsiprymnus, equinus, fulvorufulva, gazella, gnou, leche, marsupialis, niger, oreotragus, oryx, ourebi, scriptus, strepsiceros, taurinus).  Six tooth types were considered here lower/mandibular molars 1, 2, and 3 (LM1, LM2, LM3) and upper/maxillary molars 1, 2, 3 (UM1, UM2, UM3).  Specific counts of the sample sizes of each tooth type and tribe can be found in table \ref{sampsize}.

For each tooth specimen in a specific tooth type (each tooth type was considered separately), the raw representation of each shape was comprised of 60 points traced around the occlusal surface of the tooth.  For each tooth, these 60 points were split into two groups roughly divided by a line connecting the mesostyle to the entostyle (metastylid to the ectostylid) of mandibular teeth (maxillary teeth) (Full deatils in appenidix).  An example of this can be seen in Figure \ref{ToothCutter}.  This type of cut was chosen for this simulation study as this is a break point commonly observed by biological anthropologists in fossilized bovid teeth.

%Might switch $k$ to $l$ because of k-nearest neighbors. 
Then, for each half of a tooth specimen, the donee was matched against all the remaining complete teeth (partial teeth were not matched against the full tooth that the partial tooth was derived from) in that specific tooth type to find a set of $k$ suitable donor teeth. The donee was then completed $M$ times by randomly drawing $M$ teeth with replacement from the set of $k$ donor teeth.  The selected donor teeth were then used to to complete the tooth as described in section XXX.

The resulting imputed teeth were then used to evaluate the performance of several different types of analyses.  Classification is of major interest to biological anthropologists, and a k-nearest neighbors approach to classification was evaluated as well as a (kernel based approach using Frechet means was evaluated maybe?).   For k-nearest neighbors, both partial matching and matching after imputation was checked.

In additional, the total area of the tooth is of interest to biological anthropologists.  The accuracy of total area estimation using imputations was also evaluated to test if the total area of a partially observed tooth can be accurately estiamted using this imputation procedure.

We checked different values of M and k and included scaling vs not sclaing.

@@

complete_shape <- t(ptsTrainList[[1]][[1]])
#plot(t(complete_shape), type = "l")
#points(t(complete_shape), pch = 16)

@@

side <- 2
partial_shape <- t(tooth_cutter(ptsTrainList[[tooth]][[1]])[[side]])
points(t(partial_shape), pch = 16, col = "blue")
points(t(partial_shape), type = "l", col = "blue")
@
\caption{Example cuts for an Alcelaphini LM1 (DSCN2871)}
\label{ToothCutter}
\end{figure}

<<>>=
  k = 5
  M = 5
  side <- 1
  tooth <- "LM1"
load(paste0("/Users/gregorymatthews/Dropbox/shapeanalysisgit/results/results20190610_side=",side,"_k=",k,"_M=",M,"_tooth=",tooth,".RData"))

plot(t(results_list[[1]]$imputed_partial$imputed[[1]]),col = "red", type = "l", xlim = c(-200,550), ylim = c(-200,200), xlab = "", ylab = "")
points(t(results_list[[1]]$imputed_partial$partial_obs))
points(t(results_list[[1]]$imputed_partial$imputed[[2]]),col = "orange", type = "l")
points(t(results_list[[1]]$imputed_partial$imputed[[3]]),col = "gold", type = "l")
points(t(results_list[[1]]$imputed_partial$imputed[[4]]),col = "darkgreen", type = "l")
points(t(results_list[[1]]$imputed_partial$imputed[[5]]),col = "blue", type = "l")
@

<<echo = FALSE>>=
# greg <- c()
# for (i in 1:length(results_list))
#   {
#   greg[i] <- results_list[[i]]$truth$Tribe
#   }
#   table(greg)
#   sum(table(greg))
#   
#   species <- c()
# for (i in 1:length(results_list))
#   {
#   species[i] <- results_list[[i]]$truth$Species
#   }
#   
#   
#   table(greg,species)
@

  \begin{table}[h!]
  \small
  \begin{center}
\begin{tabular}{ | c | c c c c c c c | c | }
  & Alcelaphini  & Antilopini & Bovini & Hippotragini & Neotragini & Reduncini & Tragelaphini & Total \\ 
  \hline
 LM1 & 106 & 27 & 22 & 26 & 45 & 76 & 53 & 355 \\ 
 LM2 & 117 & 30 & 25 & 37 & 55 & 90 & 53 & 407\\ 
 LM3 & 117 & 30 & 19 & 34 & 53 & 96 & 62 & 411 \\ 
 UM1 & 117 & 30 & 23 & 37 & 43 & 80 & 75 & 405 \\ 
 UM2 & 118 & 30 & 23 & 41 & 53 & 97 & 94 & 456\\ 
 UM3 & 71 & 30 & 17 & 58 & 52 & 110 & 78 & 416\\ 
 \hline
\end{tabular}
\end{center}
\caption{Sample sizes for the simulation study by tooth type and tribe}
\label{sampsize}
\end{table}

@@

\subsection{Numerical Results}

<<>>=
  for (tooth in c("LM1","LM2","LM3","UM1","UM2","UM3")){
  for (side in 1:2){print(c(tooth, side))
load(paste0("/Users/gregorymatthews/Dropbox/shapeanalysisgit/results/summaries/results20190610_side=",side,"_k=5_M=5_tooth=",tooth,"_summaries.RData"))

}}
@
